For finitary induction-Induction, Induction is Enough

### 2020-04-09 paper/paper.tex
other fixes

\begin{abstract}
  Inductive-inductive types (IITs) are a generalisation of inductive types in
  type theory. They allow the mutual definition of types with multiple sorts
  where later sorts can be indexed by previous ones. An example is the James
  Chapman-style syntax of type theory with conversion relations for each sort
  where e.g.\ the sort of types is indexed by contexts. It follows from previous
  work that all finitary IITs can be constructed from a quotient
  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
  a small domain-specific type theory where a context is a signature for an
  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
  thereby showing a reduction of all finitary IITs to inductive types.  We rely
  on function extensionality and uniqueness of identity proofs. We formalised
  the construction of the theory of IIT signatures in Agda.
\end{abstract}

@@

Many mutual inductive types can be reduced to indexed inductive types, where the
index disambiguates different sorts. An example is the predicates (or \emph{indexed sorts})
$\mathsf{isEven}$ and $\mathsf{isOdd}$ defined mutually by the following
constructors.
\begin{alignat*}{5}
  & \mathsf{isEven} && : \N\ra\Set \\
  & \mathsf{isOdd} && : \N\ra \Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,(n+1) \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,(n+1)
\end{alignat*}
These can be reduced to the following single inductive family.
\begin{alignat*}{5}
  & \mathsf{isEven?} && : \mathsf{Bool}\ra\N\ra\Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven?}\,\mathsf{true}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,(n+1) \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,(n+1)
\end{alignat*}

@@

The elimination principle for the above IIT has two motives and one method
for each constructor. It is an example of a \emph{recursive-recursive}
function (following the nomenclature of \cite{forsberg-phd}) which
means that there are two mutually defined functions, where the type of
the second function depends on the first function. The proof assistant
Agda \cite{norell07thesis} allows the definition of such functions
(even from non-IITs) and it is also currently the only proof assistant
supporting IITs\footnote{An experimental version of Coq with IITs is also
  available on github.}. The elimination principle for $\Con$--$\Ty$ says that
given the following components,
where the variable $\Gamma$ is implicitly universally quantified when required
(we indeed sometimes omit universal quantification in the following, for readability),
\begin{alignat*}{5}
  & \Con^D && : \Con\ra\Set \\
  & \Ty^D && : \Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & \mathsf{empty}^D && : \Con^D\,\mathsf{empty} \\
  & \mathsf{ext}^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,A\ra \Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
  & \U^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
  & \El^D && : (\Gamma^D:\Con^D\,\Gamma)\ra\,\Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,(\U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
%  & Pi^D && : (\Gamma^D:\Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*}
we obtain two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\Gamma:\Con)\ra \Con^D\,\Gamma \\
  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra \Ty^D\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = \mathsf{empty}^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = \mathsf{ext}^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = \U^D\,(\mathsf{elimCon}\,\Gamma) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = \El^D\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}

@@

The theory of signatures in \cite{Kaposi:2019:CQI:3302515.3290315} is a small
type theory in which every context encodes a valid quotient inductive-inductive
type (QIIT) signature. For example, the signature for natural numbers is simply
the context $(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$
and $suc$ are variable names). \cite{Kaposi:2019:CQI:3302515.3290315} showed
that if this theory of signatures exists, then any finitary QIIT can be derived
from it via a term model construction. If we omit the type former for equality
constructors from the theory of signatures, we get a theory of finitary IIT
signatures, and the construction of \cite{Kaposi:2019:CQI:3302515.3290315} still
works: constructing an arbitrary IIT from the theory of IIT signatures does not
need the $\mathsf{Id}$ type former to be present.

However, the QIIT term model construction assumes that we already have the
syntax of QIIT signatures, and we can do induction over it. Hence, it only shows
that all finitary QIITs are derivable from a particular fairly complicated QIIT.
Likewise, restriced to finitary IITs, we only know that they are constructible
from a particular QIIT.

>>>>

\begin{abstract}
  Inductive-inductive types (IITs) are a generalisation of inductive types in
  type theory. They allow the mutual definition of types with multiple sorts
  where later sorts can be indexed by previous ones. An example is the James
  Chapman-style syntax of type theory with conversion relations for each sort
  where e.g.\ the sort of types is indexed by contexts. It follows from previous
  work by T. Altenkirch, A. Kaposi, and A. Kov{\'a}cs, that all finitary IITs can be constructed from a quotient
  inductive-inductive type (QIIT), namely the theory of IIT signatures. This is
  a small domain-specific type theory where a context is a signature for an
  IIT. In this paper we construct the theory of IIT signatures using only inductive types,
  thereby showing a reduction of all finitary IITs to inductive types.  We rely
  on function extensionality and uniqueness of identity proofs. We formalised
  the construction of the theory of IIT signatures in Agda.
\end{abstract}

@@

Many mutual inductive types can be reduced to indexed inductive types, where the
index disambiguates different sorts. An example is the mutual inductive datatype
with two sorts
$\mathsf{isEven}$ and $\mathsf{isOdd}$, defined by the following
constructors.
\begin{alignat*}{5}
  & \mathsf{isEven} && : \N\ra\Set \\
  & \mathsf{isOdd} && : \N\ra \Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isOdd}\,n\ra\mathsf{isEven}\,(n+1) \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven}\,n\ra\mathsf{isOdd}\,(n+1)
\end{alignat*}
These can be reduced to the following single inductive family.
\begin{alignat*}{5}
  & \mathsf{isEven?} && : \mathsf{Bool}\ra\N\ra\Set \\
  & \mathsf{zeroEven} && : \mathsf{isEven?}\,\mathsf{true}\,\mathsf{zero} \\
  & \mathsf{sucEven} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{false}\,n\ra\mathsf{isEven?}\,\mathsf{true}\,(n+1) \\
  & \mathsf{sucOdd} && : (n:\N)\ra\mathsf{isEven?}\,\mathsf{true}\,n\ra\mathsf{isEven?}\,\mathsf{false}\,(n+1)
\end{alignat*}

@@

The elimination principle for the above IIT has two motives and one method
for each constructor. It is an example of a \emph{recursive-recursive}
function (following the nomenclature of \cite{forsberg-phd}) which
means that there are two mutually defined functions, where the type of
the second function depends on the first function. The proof assistant
Agda \cite{norell07thesis} allows the definition of such functions
(even from non-IITs) and it is also currently the only proof assistant
supporting IITs\footnote{An experimental version of Coq with IITs is also
  available on github.}.

We now state  
the elimination principle for $\Con$--$\Ty$.
We implicitly universally quantify over the variables $\Gamma$ of type $\Con$
and $A$ of type $\Ty$ when required (later in the development, we may similarly omit universal quantification).
Given the following components,
\begin{alignat*}{5}
  & \Con^D && : \Con\ra\Set \\
  & \Ty^D && : \Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & \mathsf{empty}^D && : \Con^D\,\mathsf{empty} \\
  & \mathsf{ext}^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,A\ra \Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
  & \U^D && : (\Gamma^D:\Con^D\,\Gamma)\ra \Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
  & \El^D && : (\Gamma^D:\Con^D\,\Gamma)\ra\,\Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,(\U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
%  & Pi^D && : (\Gamma^D:\Con^D\,\Gamma)(A^D:\Ty^D\,\Gamma^D\,A)\ra \Ty^D\,(\mathsf{ext}^D\,\Gamma^D\,A^D)\ra\,\Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*}
we obtain two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\Gamma:\Con)\ra \Con^D\,\Gamma \\
  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra \Ty^D\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = \mathsf{empty}^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = \mathsf{ext}^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = \U^D\,(\mathsf{elimCon}\,\Gamma) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = \El^D\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}

@@

The theory of signatures in \cite{Kaposi:2019:CQI:3302515.3290315} is a small
type theory in which every context encodes a valid quotient inductive-inductive
type (QIIT) signature. For example, the signature for natural numbers is simply
the context $(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of length three ($Nat$, $zero$
and $suc$ are variable names). \cite{Kaposi:2019:CQI:3302515.3290315} showed
that if this theory of signatures exists, then any finitary QIIT can be derived
from it via a term model construction.
Here, by finitary QIIT, we mean that infinitary constructors, such as the following
one, are not allowed:
\[ TODO \]
If we omit the type former for equality
constructors from the theory of signatures, we get a theory of finitary IIT
signatures, and the construction of \cite{Kaposi:2019:CQI:3302515.3290315} still
works: constructing an arbitrary IIT from the theory of IIT signatures does not
need the $\mathsf{Id}$ type former to be present.

However, the QIIT term model construction assumes that we already have the
syntax of QIIT signatures, and we can do induction over it. Hence, it only shows
that all finitary QIITs are derivable from a particular fairly complicated QIIT.
Likewise, restricted to finitary IITs, we only know that they are constructible
from a particular QIIT.

### 2020-04-09 paper/universaliit.tex
other fixes

\subsubsection{Untyped syntax}
The untyped syntax is defined as the following inductive datatype. Variables are
modeled as de Bruijn indices, i.e.\ as natural numbers pointing to a position in
the context. We use the additional default constructor $\erru:\Tmu$ in case of
error (ill-scoped substitution). The typing judgments will not mention $\erru$.
The main interest of $\erru$ is that it behaves like a closed term (which the
theory of signatures lacks), in the sense that it is
invariant under substitution. This makes expected equalities about 
substitution (see Section~\ref{ss:exch-laws}) true even in the ill-typed case, thus reducing the amount of hypotheses.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4-5) Metatheoretic parameters} \\[0.5em]
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(6) Default value} \\[0.5em]
  & \erru && : \Tmu
\end{alignat*}

>>>>

\subsubsection{Untyped syntax}
The untyped syntax is defined as the following inductive datatype. Variables are
modeled as de Bruijn indices, i.e.\ as natural numbers pointing to a position in
the context. We use the additional default constructor $\erru:\Tmu$ in case of
error (ill-scoped substitution). The typing judgments will not mention $\erru$.
The main interest of $\erru$ is that it behaves like a closed term (which the
theory of signatures lacks), in the sense that it is
invariant under substitution. This makes expected equalities about 
substitution true even in the ill-typed case,
thus reducing the number of hypotheses for the corresponding lemmas
(see Section~\ref{ss:exch-laws}).
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4-5) Metatheoretic parameters} \\[0.5em]
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(6) Default value} \\[0.5em]
  & \erru && : \Tmu
\end{alignat*}
